openTSNE: a modular Python library for t-SNE dimensionality reduction and embedding

### 2020-07-06 paper/policar2020-opentsne.tex
Add initial draft

%%% Put your definitions there:
\startlocaldefs
\endlocaldefs

@@

\begin{abstract}
Point-based visualisations of large, multi-dimensional data from
molecular biology can reveal meaningful clusters. One of the
most popular techniques to construct such visualisations is t-distributed
stochastic neighbor embedding (t-SNE), for which a number of extensions have
recently been proposed to address issues of scalability and the quality of the resulting
visualisations. We introduce openTSNE, a modular Python library that implements
the core t-SNE algorithm and its extensions. The library is orders of magnitude
faster than existing popular implementations, including those from
scikit-learn. Unique to openTSNE is also the mapping of new data to existing embeddings, which
can surprisingly assist in solving batch effects.
\end{abstract}

@@

\begin{keyword}
\kwd{sample}
\kwd{article}
\kwd{author}
\end{keyword}

@@

The abundance of high-dimensional data sets in molecular biology calls for
techniques for dimensionality reduction, and in particular for methods that can
help in the construction of data visualizations. Popular approaches for
dimensionality reduction include principal component analysis, multidimensional
scaling, and uniform manifold approximation and projections~\cite{2018arXivUMAP}. Among
these, t-distributed stochastic neighbor embedding
(t-SNE)~\cite{maaten2008visualizing} lately received much attention as it can
address high volumes of data and reveal meaningful clustering structure. Most
of the recent reports on single-cell gene expression data start with an
overview of the cell landscape, where t-SNE embeds high-dimensional expression
profiles into a two-dimensional space~\cite{macosko2015highly,cao2019single,tasic2018shared}.
Fig.~\ref{fig:tsne}.a presents an example of one such embedding.

Despite its utility, t-SNE has been criticized for poor scalability
when addressing large data sets, lack of global organization \em t-SNE
focuses on local clusters that are arbitrarily scattered in the
low-dimensional space \em and absence of theoretically-founded implementations to
map new data into existing embeddings~\cite{ding2018interpretable,becht2019dimensionality}. Most of these shortcomings, have
recently been addressed.
\cite{fi_tsne} sped-up the method through an interpolation-based
approximation, reducing the time complexity to be merely linear dependent on the number of samples. \cite{art_of_using_tsne} proposed
several techniques to improve global positioning, including estimating similarities with a mixture of
Gaussian kernels. While no current popular
software library supports mapping of new data into reference embedding,
\cite{parametric_tsne} proposed a related approach
using neural networks.

%% TODO: END FROM FIRST PAPER

\begin{figure*}[htbp]
  \includegraphics[width=\textwidth]{macosko2015}
  \caption{\label{fig:macosko}}
\end{figure*}

@@

We introduce openTSNE, a comprehensive Python library that
implements t-SNE and all its recently proposed extensions. The library is
compatible with the Python data science ecosystem (e.g., {\tt numpy}, {\tt
sklearn}, {\tt scanpy}). Its modular design encourages extendibility and
experimentation with various settings and changes in the analysis pipeline.

@@

The following code is used to generate Fig.~\ref{fig:tasic}
\begin{minted}[fontsize=\footnotesize]{python}
# Perplexity 30 (standard t-SNE)
perp_30 = openTSNE.TSNE(
    perplexity=30, metric="cosine"
).fit(X)

# Perplexity 500 (emphasise global structure)
perp_500 = openTSNE.TSNE(
    perplexity=500, metric="cosine"
).fit(X)

# Multiscale kernel with perplexities 30,500
affinities = openTSNE.affinity.Multiscale(
    X, perplexities=[30, 500], metric="cosine"
)
init = openTSNE.initialization.pca(X)
ms = openTSNE.TSNEEmbedding(init, affinities)
ms = ms.optimize(n_iter=250, exaggeration=12)
ms = ms.optimize(n_iter=250, exaggeration=1)

# Expose finer structure using smaller dof
perp_30 = openTSNE.TSNE(
    perplexity=30, metric="cosine", dof=0.6
).fit(X)
\end{minted}

@@

The following code is used to generate Fig.~\ref{fig:cao}
\begin{minted}[fontsize=\footnotesize]{python}
indices = np.random.permutation(list(range(x.shape[0])))
reverse = np.argsort(indices)
# Split the data into a sample/remainder sets
x_sample, x_rest = x[indices[:25000]], x[indices[25000:]]
y_sample, y_rest = y[indices[:25000]], y[indices[25000:]]
# Create an initial embedding using a subset of the data
sample_embedding = openTSNE.TSNE(
    initialization="spectral",
    perplexity=500,
    metric="cosine",
).fit(x_sample)
# Position the remainder of the points to their nearest
# neighbor in the sample embedding
rest_init = sample_embedding.prepare_partial(x_rest, k=1)
init_full = np.vstack(
    [sample_embedding, rest_init]
)[reverse]
init_full = openTSNE.initialization.rescale(init_full)
# Calculate the affinity model using all the data points
affinities = openTSNE.affinity.PerplexityBasedNN(
    x, perplexity=30, metric="cosine"
)

embedding = openTSNE.TSNEEmbedding(
    init_full, affinities
%)
exag12 = embedding.optimize(n_iter=500, exaggeration=12)
exag4 = exag12.optimize(n_iter=500, exaggeration=4)
exag2 = exag4.optimize(n_iter=500, exaggeration=2)
exag1 = exag2.optimize(n_iter=500, exaggeration=1)
\end{minted}

@@

The following code is used to generate Fig.~\ref{fig:transform}
\begin{minted}[fontsize=\footnotesize]{python}
# Assuming we have prepared a reference embedding and
# assuming we have aligned new_X to the reference X matrix
new_embedding = reference_embedding.transform(new_X)
\end{minted}

We want to mention that this works for mitigating batch effects in our earlier work~\cite{policar2019embedding}.

@@

\section*{Discussion}
Text for this section \ldots

@@

Widespread adoption of novel techniques is almost universally accompanied by easy-to-use and easy-to-install software. While the t-SNE implementation in scikit-learn and Rtsne fits this requirement, it implements the outdated Barnes-Hut approximation scheme which is unsuitable for larger data sets containing up to millions of data points. Other C++ implementations such as MulticoreTSNE and FIt-SNE are more suitable for larger data sets, but do not provide precompiled binaries to their software, requiring users to compile the software themselves. This is problematic from a usability standpoint, as many users often either do not know how to compile C++ source code or use Windows, which does not include a C++ compiler by default.

openTSNE is a Python package which focuses on ease-of-use and extensibility. We provide precompiled binaries for all major Python version on all major platforms, making the installation process as seamless as possible. openTSNE can be installed through the Python Package Index (PyPI) or through conda on the conda-forge channel. openTSNE is designed to be familliar to Python users and loosely follows the scikit-learn API, which is well established in the Python data-science ecosystem. openTSNE implements both the Barnes-Hut and FIt-SNE approximation schemes, allowing it to be used on data sets containing millions of data points. While the Python virtual machine introduces some overhead, the runtime is comparable to that of its pure C++ counterpart -- FIt-SNE. Finally, openTSNE is extensible. Its modular interface enables researchers to quickly experiment with different settings and easily incorporate their own components into the software.

\begin{description}
\item[Alternative initialization schemes] Spectral and PCA, cite dmitry and george paper

\item[Incorporates latest developments] $lr=N/12, n_iter=750$

\end{description}

@@

\section*{Availability}
Text for this section \ldots

@@

\section*{Acknowledgements}
  Text for this section \ldots
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                  The Bibliography                       %%
%%                                                         %%
%%  Bmc_mathpys.bst  will be used to                       %%
%%  create a .BBL file for submission.                     %%
%%  After submission of the .TEX file,                     %%
%%  you will be prompted to submit your .BBL file.         %%
%%                                                         %%
%%                                                         %%
%%  Note that the displayed Bibliography will not          %%
%%  necessarily be rendered by Latex exactly as specified  %%
%%  in the online Instructions for Authors.                %%
%%                                                         %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

%%% Put your definitions there:
\startlocaldefs
\newcommand{\opentsne}{\textsf{openTSNE}}
\endlocaldefs

@@

\begin{abstract}
Point-based visualisations of large, multi-dimensional data from molecular biology can reveal meaningful clusters. One of the most popular techniques to construct such visualisations is t-distributed stochastic neighbor embedding (t-SNE), for which a number of extensions have recently been proposed to address issues of scalability and the quality of the resulting visualisations. We introduce \opentsne, a modular Python library that implements the core t-SNE algorithm and its extensions. The library is orders of magnitude faster than existing popular implementations, including those from scikit-learn. Unique to \opentsne\ is also the mapping of new data to existing embeddings, which can surprisingly assist in solving batch effects.
\end{abstract}

@@

\begin{keyword}
\kwd{t-SNE}
\kwd{embedding}
\kwd{visualization}
\kwd{dimensionality reduction}
\end{keyword}

@@

The abundance of high-dimensional data sets in molecular biology calls for
techniques for dimensionality reduction, and in particular for methods that can
help in the construction of data visualizations. Popular approaches for
dimensionality reduction include principal component analysis (PCA), multidimensional
scaling, t-distributed stochastic
neighbor embedding (t-SNE)~\cite{maaten2008visualizing}, and uniform manifold approximation and
projections (UMAP)~\cite{2018arXivUMAP}. Among these, t-SNE lately received much
attention as it can address high volumes of data and reveal meaningful
clustering structure. Most of the recent reports on single-cell gene expression
data start with an overview of the cell landscape, where t-SNE embeds
high-dimensional expression profiles into a two-dimensional
space~\cite{macosko2015highly,cao2019single,tasic2018shared}.
Figs.~\ref{fig:macosko}.a and \ref{fig:macosko}.b are two examples of such
embeddings.

\begin{figure*}[htbp]
  \includegraphics[width=\textwidth]{macosko2015}
  \caption{\label{fig:macosko}The figure shows the progression of t-SNE embeddings following theoretical advances, which are all included in \opentsne. The data in \textbf{(a)} and \textbf{(b)} represent 44,808 single-cell gene-expression profiles of mouse retinal cells from the study by Macosko \textit{et al.}~\cite{macosko2015highly}. Similarly, the data in \textbf{(c)} contains 27,499 expression profiles from mouse retinal cells from Shekhar \textit{et al.}~\cite{shekhar2016comprehensive}. \textbf{(a)} We construct a t-SNE embedding following the parameter choices from the original publication from Maaten \& Hinton~\cite{maaten2008visualizing}. The visualization shows poor preservation of the global organisation of clusters, resulting from random initialization and an affinity model focused on preserving local neighborhoods. \textbf{(b)} shows a modern t-SNE embedding, utilizing the latest theoretical advances and practical recommendations. The embedding is constructed using a multi-scale affinity model, preserving both short-range and long-range interactions between data points, and is initialized such that the global layout is meaningful. Unlike in \textbf{(a)}, the green and blue clusters representing different sub-types of amacrine and bipolar cells are now localized to the same regions of the space, indicating a higher level of similarity than to other cell types. \textbf{(c)} shows how existing t-SNE reference atlases can be used to embed new samples into existing embeddings. The positions of new data points correspond to cell types from the reference atlas.}
\end{figure*}

Despite its utility, t-SNE has often been criticized for poor scalability when
addressing large data sets, lack of global organization -- t-SNE focuses on local
clusters that are arbitrarily scattered in the low-dimensional space  -- and
absence of theoretically-founded implementations to map new data into existing
embeddings~\cite{ding2018interpretable,becht2019dimensionality}. Most of these
shortcomings, have recently been addressed. Linderman \textit{et
al.} sped-up the method through an interpolation-based
approximation, reducing the time complexity to be merely linear dependent on
the number of samples~\cite{linderman2019fast}. Kobak \textit{et al.} proposed
several techniques to improve global positioning, including estimating
similarities with a mixture of Gaussian kernels~\cite{kobak2019art}. In our previous work, we presented
a principled approach for embedding new samples into existing visualizations~\cite{policar2019embedding}.

@@

We introduce \opentsne, a comprehensive Python library that implements t-SNE and all its recently proposed extensions. Most notable among these are the implementation of efficient approximation schemes~\cite{van2014accelerating,linderman2019fast} allowing the embedding of millions of data points, the addition of new samples into existing t-SNE embeddings~\cite{policar2019embedding}, better initialization schemes~\cite{kobak2019umap} leading to more globally consistent layouts, variable degrees of freedom~\cite{kobak2019heavy} allowing the inspection of data at different levels of resolution, multiscale similarity kernels~\cite{kobak2019art} which emphasize the preservation of both short-range and long-range interactions, and improved default parameters~\cite{belkina2019automated} which produce better embeddings at lower computational cost. The library is compatible with the Python data science ecosystem (e.g., \textsf{numpy}, \textsf{scikit-learn}, \textsf{scanpy}) providing a familiar and intuitive user interface to new users. Its modular design encourages extensibility and experimentation with various settings and changes in the analysis pipeline.

\opentsne\ is open-source and has become an integral part of the Python data-science ecosystem. The package has received over 600 Github stars, and has been downloaded hundreds of thousands of times, averaging in thousands of weekly downloads. The source code is publicly available on Github at \url{https://github.com/pavlin-policar/openTSNE}, and can be installed from \textsf{PyPI} and \textsf{conda-forge} -- the two most widely adopted Python package managers.

Accessibility is one of the core design principles of \opentsne. Our aim is to make the latest theoretical advances  in t-SNE as accessible as possible to as many users as possible. As such, our API is based on the familiar scikit-learn interface~\cite{sklearn_api}, enabling a smooth transition from other Python data science libraries. For example, Fig.~\ref{fig:tasic} depicts four different t-SNE embeddings, each based on recent advances in t-SNE visualizations. The code used to generate these embeddings is listed below.

\begin{minted}[
fontsize=\footnotesize,
frame=lines,
framesep=2mm,
baselinestretch=1.1
]{python}
import openTSNE
from openTSNE.affinity import Multiscale
# a - standard t-SNE
tsne1 = openTSNE.TSNE(perplexity=30).fit(X)
# b - high perplexity  for globa structure
tsne2 = openTSNE.TSNE(perplexity=500).fit(X)
# c - multiscale kernel for local/global structure
aff = Multiscale(X, perplexities=[30, 500])
tsne3 = openTSNE.TSNE(affinities=aff).fit(X)
# d - decrease dof for higher resolution
tsne4 = openTSNE.TSNE(dof=0.6).fit(X)
\end{minted}

\noindent The code snippet takes as input a \textsf{numpy} array or \textsf{scipy} sparse matrix \texttt{X} and produces four \texttt{TSNEEmbedding} objects. These may subsequently be subject to further optimization under different parameter settings, or may be used to embed new samples into the embedding landscape. Additional parameters are omitted for clarity, but can be found in accompanying notebooks publicly available at \url{https://github.com/pavlin-policar/opentsne-paper}.

\section*{Discussion}

\subsection*{Uncovering Structure in High-Dimensional Data Sets}

Dimensionality reduction techniques implicitly assume that high-dimensional data lies on a lower-dimensional manifold, which can accurately be captured by a smaller number of dimensions. However, there is no evidence to suggest that any data set can accurately be described using only two dimensions, and that any projection onto a two-dimensional plane will inevitably lead to loss of information. Therefore, it is beneficial to examine multiple embeddings, each of which provide a different perspective into the topology and other characteristics of the data. For instance, in Fig.~\ref{fig:tasic}, we show four embeddings of the same gene-expression data from the mouse brain~\cite{tasic2018shared}.

@@

Fig.~\ref{fig:tasic}.a shows a t-SNE embedding using default parameters, which clearly uncovers numerous clusters. Warm colors correspond to excitatory neurons, while cool colors represent inhibitory neurons. While clusters of excitatory and inhibitory neurons appear close together, all clusters appear somewhat equidistant from their neighobrs, and the overall relations between clusters are hidden from the user. On the other hand, Fig.~\ref{fig:tasic}.b focuses on preserving larger neighborhoods of points, resulting in a much more globally consistent layouts where relations between clusters become more apparent. In this embedding, it is evident that there is one large group of excitatory neurons, and two related groups of inhibitory neurons. Unfortunately focusing on preserving large neighborhoods often results in smaller clusters being absorbed into larger ones. Alternatively, Fig.~\ref{fig:tasic}.c uses multi-scale similarity kernels which aim to preserve both the global organisation of clusters as well as prevent smaller clusters from being obscured by larger ones. Fig.~\ref{fig:tasic}.d shows another lever available to users used to uncover clustering structure at different levels of resolution. Kobak \textit{et al.}~\cite{kobak2019heavy} found that decreasing the degrees of freedom in the t-distribution an reveal smaller, more compact clusters, otherwise easily missed in other embeddings.

When faced with a data set containing millions of data points, standard t-SNE embeddings often become unwieldy, with blurred cluster boundaries, large clusters absorbing smaller ones, and relationships between clusters becoming increasingly difficult to interpret. Fig.~\ref{fig:cao}.a contains expression profiles of 2,058,652 single cells captured at different time points in mouse development. The embedding clearly indicates numerous clusters with some transition between time points, but this is largely difficult to reason about. Kobak \& Berens observed that increasing attractive forces between similar data points controlled via the \textit{exaggeration} parameter leads to more compact clusters, which leads to more informative visualizations~\cite{kobak2019art}. For instance, \ref{fig:cao}.b increases the doubles the default exaggeration rate and begins to reveal the overall structure of the data. Doubling the exaggeration again in \ref{fig:cao}.c, we can now easily observe that the data is comprised of two large clusters, and eight smaller ones. Coincidentally, embeddings with this level of exaggeration roughly correspond to embeddings produced by UMAP~\cite{TODO}.

@@

Exaggeration can be used to highlight transitions between cell-states in developmental studies. Standard t-SNE often produces embeddings with clearly-defined, discrete clusters. We have shown that users have access to several parameters to control the granularity of the clustering structure. However, such properties are undesired in developmental studies as we assume that the cells follow a continuous path between cell states. To this end, other embedding techniques such as UMAP and ForceAtlas2 are often used, as they tend to better capture the continuity between cell states. Recently, Kobak \textit{et al.}~\cite{TODO} have shown that embeddings produced by t-SNE with exaggeration values of 4 and $\sim30$ produce embeddings which are visually very similar to UMAP and ForceAtlas2. For example, in Fis.~\ref{fig:cao}.a the colors clearly indicate some transitional states, which are made much more apparent in Fig.~\ref{fig:cao}.c.

Unlike other popular dimensionality reduction methods such as principal component analysis or autoencoders, t-SNE is a non-parametric method and does not define an explicit mapping to the embedding space. Therefore embeddings of new data points need to be found through the use of optimization techniques~\cite{policar2019embedding}. \opentsne\ is currently the only publicly available library allowing users to add new samples to existing embeddings.

Figs.~\ref{fig:macosko}.b and \ref{fig:macosko}.c demonstrate how we can use a previously annotated single-cell data set and embed data from a different single-cell experiment into the reference landscape. The reference data taken from Macosko \textit{et al.}~\cite{macosko2015highly} contains gene-expression profiles from mouse retinal cells. Taking data from and experiment focusing only on bipolar retinal cells by Shekhar \textit{et al.} and embedding these cells into the reference embedding, we can see that the majority of new cells correctly map to the bipolar cell clusters in the reference.

Embedding single cells into existing reference atlases can also be useful for cell-type classification in cases where cell identities are not yet known. For instance, in Fig.~\ref{fig:transform}, we construct a reference embedding using data from Hochgerner \textit{et al.}~\cite{hochgerner2018conserved} containing  gene-expression profiles of cells from the mouse brain. The authors provide cell annotations for each cell, and we can verify their accuracy by visualizing well-established gene markers for the major cell types. We can then embed similar data from Harris \textit{et al.}, where annotations are provided only for neuronal cells, and quickly identify other non-neuronal cell types, including oligodendrocytes and astrocytes. We then use the marker genes to validate that new cells were correctly mapped into the reference landscape.

@@

These two examples demonstrate how \opentsne\ can be used to quickly gain insight into newly-sequenced, single-cell data sets by utilizing already existing, publicly-available, reference atlases. This approach is extremely general and is not limited to single-cell gene-expression data, but can be applied to any tabular data.

\subsection*{Versatility}

Versatility is another of  \opentsne s core design principles. Kobak \& Berens recently introduced several recommendations and tricks to obtain better and more meaningful t-SNE embeddings~\cite{kobak2019art}. These include the use of multi-scale similarity kernels, slowly easing the perplexity parameter, and increasing exaggeration when working with larger data sets. \opentsne\ provides an extremely flexible API, where these improvements can be incorporated in a few lines of code. Furthermore, \opentsne\ additionally supports custom affinity models, e.g. multi-scale Gaussian mixtures or uniform kernels, allowing for quick experimentation. This also enables users to construct t-SNE embeddings on non-tabular data, e.g. relational data, as the only affinity-model requirement is some notion of similarity between data points. Finally, our comprehensive callback system can utilized to monitor and adapt different stages of the optimization phase and has been used to construct visually appealing animations of the t-SNE optimization process.

\subsection*{Speed}

One of t-SNE's common criticisms is poor scalability to large data sets containing up to millions of data points~\cite{becht2019dimensionality}. These criticisms are largely based on two factors. Until quite recently, most popular implementations of t-SNE were based on the Barnes-Hut approximation scheme developed by van der Maaten in 2014~\cite{van2014accelerating} with asymptotic complexity $\mathcal{O}(N \log N)$. More importantly, this misconception was exacerbated further by the fact that the most widely-used implementation of t-SNE came from \textsf{scikit-learn}, which exhibits very poor runtime performance compared to its C++ counterpart -- \textsf{MulticoreTSNE}. Even with the Barnes-Hut approximation, the multi-threaded \textsf{MulticoreTSNE} implementation~\cite{Ulyanov2016} is perfectly capable of creating t-SNE embeddings of millions of data points in a matter of hours on widely-accessible, consumer-grade CPUs (Fig.~\ref{fig:benchmarks}). Recently, Linderman \textit{et al.} developed a new approximation scheme -- FIt-SNE -- which further reduces the asymptotic time complexity to $\mathcal{O}(N)$, embedding large data sets in a matter of minutes.

Fig.~\ref{fig:benchmarks} shows benchmarks of four Python t-SNE implementations, including \textsf{scikit-learn} v0.23.1, \textsf{MulticoreTSNE} v0.1, \textsf{FIt-SNE} v1.1.0, and \opentsne\ v0.4.3. We perform benchmarks on two processors, one representing usage on a personal computer and one on a high-performance computing platform. The Intel(R) Core i7 is commonly found in consumer-grade laptop computers, while Intel(R) Xeon(R) processors are often found on high-performance computing machines. Benchmarks were run with the original t-SNE parameters for 1000 iterations, as some implementations do not enable their modification.

@@

Fig.~\ref{fig:benchmarks} demonstrates that both \textsf{FIt-SNE} and \opentsne\ exhibit much better runtime performance than their Barnes-Hut counterparts -- \textsf{scikit-learn} and \textsf{MulticoreTSNE}. Notice that even though \textsf{FIt-SNE} and \opentsne\ both implement the same approximation scheme, \opentsne\ tends to be slightly slower. \opentsne\ is written in Python, which inevitably introduces some runtime overhead compared to its pure C++ counterpart. However, Fig.~\ref{fig:benchmarks} demonstrates the runtime of \opentsne\ is comparable to that of \textsf{FIt-SNE} and much faster than its Barnes-Hut competitors. Additionally, modern computer processors generally contain more than a single core, allowing us to make use of multi-threading, which drastically reduces the runtime differences between \opentsne\ and \textsf{FIt-SNE}. On the server-grade processor, \opentsne\ is actually slightly faster than its pure C++ counterpart. We speculate that this is likely due to the higher level API \opentsne\ uses for linear algebra operations, which may make better use of the Intel Math Kernel Library (MKL), which are aggressively optimized on Xeon processors.

Additionally, \opentsne\ provides a flexible API, which allows splitting up the embedding-construction process and caching slow operations, enabling the user to quickly experiment with different parameter settings.

Widespread adoption of novel techniques is almost universally accompanied by easy-to-use and easy-to-install software. While the t-SNE implementation in \textsf{scikit-learn} fits this requirement, it implements the outdated Barnes-Hut approximation scheme which is unsuitable for large data sets. Other C++ implementations such as \textsf{MulticoreTSNE} and \textsf{FIt-SNE} are more suitable to larger data sets, but do not provide precompiled binaries, requiring users to compile the software themselves. This is problematic from a usability standpoint, as many users often lack the knowledge to do so. This is especially problematic with Windows users, where the operating system does not include a C++ compiler by default and correctly configuring the command line can be an arduous process.

\opentsne\ is a Python package which focuses on accessibility. We provide precompiled binaries for all major Python version on all major platforms, making the installation process as seamless as possible. \opentsne\ can be installed through the Python Package Index (PyPI) or through conda through the conda-forge channel. \opentsne\ is designed to be familiar to Python users and follows the scikit-learn API~\cite{sklearn_api}, which is well established in the Python data-science ecosystem. \opentsne\ implements multi-threaded versions of both the Barnes-Hut and FIt-SNE approximation schemes, allowing it to be used on data sets containing millions of data points. While the Python virtual machine introduces some overhead, the runtime is comparable to that of its pure C++ counterpart -- \textsf{FIt-SNE}. Finally, \opentsne\ is extensible. Its modular design enables researchers to quickly experiment with different settings and easily incorporate their own components into the software, and has already been used by various researchers in their own work.

\subsection*{Comparison to UMAP}

Uniform Manifold Approximation and Projection (UMAP)~\cite{2018arXivUMAP} has often been touted as a silver bullet to the problems listed above, exhibiting impressive runtime on large data sets and organizing resulting clusters in a coherent manner. However, most performance comparisons to t-SNE were performed against older t-SNE implementations. Furthermore, the optimization procedure of UMAP is inherently sequential and does not easily lend itself to parallelization. On the other hand, modern t-SNE approximations are highly parallelizable and easily lend themselves to utilizing multiple cores, present on almost every consumer-grade processor. The embeddings produced by UMAP, like other force-directed layout algorithms such as t-SNE and ForceAtlas2~\cite{jacomy2014forceatlas2}, are largely dependent on their initialization. Many implementations of t-SNE begin with a randomly initialized embedding, leading to a largely scattered and incoherent layout of resulting clusters. On the other hand, UMAP embeddings are initialized using Laplacian eigenmaps, a method for the preservation of global relationships between data points. Kobak \& Linderman showed in \cite{kobak2019umap} that the supposed superior preservation of global structure is largely the result of this initialization, and that t-SNE, if initialized in similar manner, performed just as well in preserving global coherence. Most modern implementation of t-SNE default to such initialization schemes, typically though principal component analysis or spectral embeddings.

We argue here not that either method is superior to the other, but rather, that they are complementary and should be used in tandem. Each may reveal different aspects of the data missed by the other.

@@

@@

\section*{Availability}

openTSNE is distributed under the BSD-3-Clause License and is publicly available as an open-source package at \url{https://github.com/pavlin-policar/openTSNE}. openTSNE is also available on PyPI and conda-forge. The data sets used and scripts used in this study are included in accompanying notebooks, publicly available at \url{https://github.com/pavlin-policar/opentsne-paper}.

@@

\section*{Acknowledgements}
We would like to thank Dmitry Kobak for various helpful discussions and best practices when using t-SNE, as well as his contributions to the source code. We would also like to thank George Linderman for his help with the FIt-SNE algorithm.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                  The Bibliography                       %%
%%                                                         %%
%%  Bmc_mathpys.bst  will be used to                       %%
%%  create a .BBL file for submission.                     %%
%%  After submission of the .TEX file,                     %%
%%  you will be prompted to submit your .BBL file.         %%
%%                                                         %%
%%                                                         %%
%%  Note that the displayed Bibliography will not          %%
%%  necessarily be rendered by Latex exactly as specified  %%
%%  in the online Instructions for Authors.                %%
%%                                                         %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
